Powering Hidden Markov Model by Neural Network based Generative Models

### 2019-08-27 manuscript/main.tex
update

Another track of related work is hybrid method of temporal neural network models and HMM. In \cite{liu2019lstmHmmHyb,buys2018bridging,vik2016rnnHmm}, a long short-term memory (LSTM) model/recurrent neural network (RNN) is combined with HMM as hybridization. A hierarchical training is carried out by: i) training a HMM first, ii) and then doing modified training of LSTM using trained HMM. This hierarchical training procedure is motivated by the intuition of using LSTM or RNN to fill in the gap where HMM can not learn.

@@

In this work, we propose a generator-mixed HMM model, termed as GenHMM, where generators are realized by neural networks to help the model gain high modeling capacity.
Our proposed model, GenHMM:
\begin{itemize}
  \item firstly has high modeling capacity of sequential data, due to the neural network based generators;
\item has advantage of easy to do training. Training of GenHMM employs expectation maximization (EM) framework. Training a GenHMM is as easy as training a GMM-HMM model, while configuration of GenHMM is flexible.
% \item has high modeling capability of sequence data, though its training and using are handy
\item is able to compute loglikelihood exactly and efficiently.
\end{itemize}
Instead of using softmax for probability representation, our GenHMM has tractability of exact loglikelihood of given sequential data, which is based on change of variable formula. To make the loglikelihood computation efficient, neural network based generators of GenHMM are implemented as flow models.

Our contributions in the paper are as follows:
\begin{itemize}
\item Proposing a neural network based HMM for sequential data modeling, i.e. GenHMM. The likelihood computation of sequential data under GenHMM can be computed exactly and efficiently.
\item Designing practical algorithm for training GenHMM under EM framework. Stochastic gradient  search in batch fashion is embedded in this algorithm.
\item Giving convergence analysis for GenHMM under the proposed learning algorithm.
\end{itemize}

\section{Generator-mixed HMM}

@@

Our framework is a Hidden Markov Model (HMM). A HMM model $\bm{H}$ defined in a hypothesis space $\Hh$, i.e. $\bm{H} \in \Hh$, is capable to model time-span signal $\ubar{\bm{x}} = \left[ \bm{x}_1, \cdots, \bm{x}_T\right]^{\intercal}$, where $\bm{x}_t\in \RR^{N}$ is the signal at time $t$, $[\cdot]^{\intercal}$ denotes transpose, and $T$ denotes the time length\footnote{The length for difference sequential data can be different.}. Using HMM for signal representation is illustrated in Figure~\ref{fig:hmm}. The model assumption is that different instant signal of $\ubar{\bm{x}}$ is generated by a different signal source associated with a hidden state of HMM, while the source selection is done by a discrete hidden markov process. We define the hypothesis set of HMM as $\Hh := \{\bm{H} | \{\Ss, \bm{q}, \bm{A}, p(\bm{x}|{s}; \bm{\Phi}_{s})\}$, where:
\begin{itemize}
\item $\Ss$ is the set of hidden states of HMM $\bm{H}$;
\item $\bm{q} = \left[ q_1, q_2, \cdots, q_{|\Ss|}\right]^\intercal$ is initial distribution of HMM $\bm{H}$ with $|\Ss|$ as cardinality of $\Ss$. For a state $s$, $q_s = p(s;\bm{H})$. In the following of this paper, we use $s_t$ to denote the state $s$ at time $t$.
\item $\bm{A}$ is the transition matrix for the HMM $\bm{H}$ of size $|\Ss| \times |\Ss|$ and $\bm{A}_{i,j} = p(s_{t+1}=j|s_{t}=i; \bm{H})$.
\item For given hidden state $s$, the density function of observable signal is $p({\bm{x}}|{s};\bm{\Phi}_{s})$, where $\bm{\Phi}_{s}$ is the parameter set that defines this probabilistic model.
\end{itemize}

In the framework of HMM, at each time instance, signal $\bm{x}_t$ is assumed to be generated by a distribution with density function $p(\bm{x}_t| s_t; \bm{\Phi}_{s_t})$, and $s_t$ is decided by the hidden markov process. This gives us the probabilistic model $p(\ubm{x};\bm{H})$.
\subsection{Generative Model of GenHMM}

@@

In this section, we introduce a neural network based generator-mixed HMM model. We term it as GenHMM. GenHMM has a hidden markov process as a typical HMM does, while the probabilistic model of GenHMM for each hidden state is induced by mixture of neural network based generators. For given state $s\in \Ss$, we define its probabilistic model in GenHMM as
\begin{equation}
  p(\bm{x}| s; \bm{\Phi}_{s}) = \sum_{\kappa=1}^{K}\pi_{s, \kappa} p(\bm{x}| s, \kappa; \bm{\theta}_{s, \kappa}),
\end{equation}
where $\kappa$ is a random variable following a categorical distribution, with probability $\pi_{s, \kappa} = p(\kappa | s; \bm{H})$. Naturally $\sum_{\kappa = 1}^{K} \pi_{s, \kappa}= 1$. Denote $\bm{\pi}_{s} = [\pi_{s,1}, \pi_{s,2}, \cdots, \pi_{s,K}]^{\intercal}$,
$\bm{\theta}_s = \left\{ \bm{\theta}_{s, \kappa}| \kappa = 1, 2, \cdots, K \right\}$. 
We define $p(\bm{x}| s, \kappa; \bm{\theta}_{s, \kappa})$ as induced distribution by a generator $\bm{g}_{s,\kappa}: \RR^{N}\rightarrow\RR^{N}$, such that $\bm{x}=\bm{g}_{s, \kappa}(\bm{z})$, where $\bm{z}$ is a latent variable following a distribution with density function $p_{s,\kappa}(\bm{z})$. By change of variable, we have
\begin{equation}\label{eq:changel-variable}
  p(\bm{x}| s, \kappa; \bm{\theta}_{s, \kappa}) = p_{s,\kappa}(\bm{z})\bigg| \det\left( \pd{\bm{g}_{s,\kappa}(\bm{z})}{\bm{z}} \right)\bigg|^{-1}.
\end{equation}

For a state $s$ of GenHMM, the induced probability distribution is illustrated in Figure~\ref{fig:gen-mix}.

@@

The first problem that we need to address is to deal with the hidden sequential variable $\ubm{s}$ and $\ubm{\kappa}$  in model $\bm{H}$. For a sequential observable variable $\ubm{x}$, $\ubm{s}$ is the hidden state sequence corresponding to $\ubm{x}$, and $\ubm{\kappa}$ is the hidden variable sequence representing the generator identity sequence that actually generates $\ubm{x}$.

@@

In this section, we detail the solution for realizing and learning GenHMM. % We begin with introducing flow model that we are going to use to model our generators $\{\bm{g}_{s, \kappa}| s\in\Ss, \kappa=1,2,\cdots, K\}$ for efficient likelihood computation. The learning method of HMM is followed then. 
The convergence analysis is discussed at the end of this section.

\subsection{Realizing $\bm{g}_{s,\kappa}$ by a Flow Model}
We use a feed-forward neural network to implement each generator $\bm{g}_{s,\kappa}$.
Define  $\bm{g}_{s,\kappa}$ as a $L$-layer neural network and we formulate its mapping by each layer's concatenation:
$\bm{g}_{s,\kappa}=\bm{g}_{s,\kappa}^{[L]}\circ \bm{g}_{s,\kappa}^{[L-1]}\circ \cdots
\circ \bm{g}_{s,\kappa}^{[1]}$, where superscipt $[l]$ denotes the $l$'s layer index and $\circ$ denotes mapping concatenation. Assume $\bm{g}_{s,\kappa}$ is invertible and denotes its inverse mapping as $\bm{f}_{s,\kappa}=\bm{g}_{s,\kappa}^{-1}$. For a latent variable $\bm{z}$ with density function $p_{s,\kappa}(\bm{z})$, the generated signal $\bm{x}$ follows an induced distribution with density function \eqref{eq:changel-variable}. We illustrate the signal flow between latent variable $\bm{z}$ and observable variable $\bm{x}$ as follows:
\begin{equation*}
  \vspace{-8pt}
  \centering
  \begin{tikzpicture}
    \node (z) at (0,0) {};
    \node at ($(z)-(0.5,0)$){$\bm{z}=\bm{h}_0$};
    \node (xi1) at (1.5,0) {$\bm{h}_1$};
    \node (xi2) at (3,0) {};
    \node (xi3) at (4.5,0){};
    \node (x) at (6,0) {};
    \node at ($(x)+(0.5,0)$){${\bm{h}_L=\bm{x}}$};
    \draw[->] ($(z) + (0.3,0.1)$) -- node[above]{$\bm{g}_{s,\kappa}^{[1]}$} ($(xi1)+(-0.3,0.1)$); 
    \draw[->] ($(xi1)-(0.3,0.1)$) -- node[below]{${\bm{f}}_{s,\kappa}^{[1]}$}($(z) - (-0.3,0.1)$);
    \draw[->] ($(xi1) + (0.3,0.1)$) -- node[above]{$\bm{g}_{s,\kappa}^{[2]}$} ($(xi2)+(-0.3,0.1)$); 
    \draw[->] ($(xi2)-(0.3,0.1)$) -- node[below]{${\bm{f}}_{s,\kappa}^{[2]}$}($(xi1) - (-0.3,0.1)$);
    \draw[->] ($(xi3) + (0.3,0.1)$) -- node[above]{$\bm{g}_{s,\kappa}^{[L]}$} ($(x)+(-0.3,0.1)$); 
    \draw[->] ($(x)-(0.3,0.1)$) -- node[below]{${\bm{f}}_{s,\kappa}^{[L]}$}($(xi3) - (-0.3,0.1)$);
    \draw[dotted,line width = 0.3 mm] (xi2) -- (xi3);
  \end{tikzpicture},
\end{equation*}
where $\bm{f}_{s,\kappa}^{[l]}$ is the $l$-th layer of $\bm{f}_{s,\kappa}$. We have $\bm{z}=\bm{f}_{s,\kappa}(\bm{x})$. If every layer of $\bm{g}_{s,\kappa}$ is invertible,
the full feed-forward neural network is invertible. Flow model, proposed in \cite{DBLP:journals/corr/DinhKB14} as an image generating model, is such a invertible
feed-forward neural network. It is further improved in subsquential works \cite{2016arXiv160508803D,kingma2018glow} for high-fidelity and high-resolution image generating and representation. Flow model also has advantages as efficient Jacobian computation and low computational complexity. %Computation of inverse mapping for flow-based network and determinant of Jacobin were addressed in the literature.

@@

We firstly address the generator learning problem, i.e. $\umax{\bm{\Theta}} \Qq(\bm{\Theta}; \bm{H}^{\mathrm{old}})$. This is boiled down to maximizing the cost function of neural networks that can be formulated as
\begin{align}\label{eq:obj-q-gen-mix}
  &\Qq(\bm{\Theta}; \bm{H}^{\mathrm{old}}) \nonumber \\
  = &\frac{1}{R}\sum_{r=1}^{R}\sum_{\ubmr{s}{r}}\sum_{\ubmr{\kappa}{r}}{p(\ubmr{s}{r}, \ubmr{\kappa}{r}| \ubmr{x}{r}; \bm{H}^{\mathrm{old}})} \sum_{t=1}^{{T}^{r}}\log\,p(\bmtr{x}{t}{r} | \smtr{s}{t}{r}, \smtr{\kappa}{t}{r}; \bm{H}) \nonumber \\
  =& \frac{1}{R}\sum_{r=1}^{R} \sum_{t=1}^{{T}^{r}} \sum_{\smtr{s}{t}{r}=1}^{|\Ss|}  \sum_{\smtr{\kappa}{t}{r}=1}^{K}p(\smtr{s}{t}{r}| \ubmr{x}{r}; \bm{H}^{\mathrm{old}})p(\smtr{\kappa}{t}{r}|\smtr{s}{t}{r}, \ubmr{x}{r}; \bm{H}^{\mathrm{old}}) \nonumber\\
  &  \log\, p(\bmtr{x}{t}{r} | \smtr{s}{t}{r}, \smtr{\kappa}{t}{r}; \bm{H}), 
\end{align}
where $T^r$ is the length of the $r$-th sequential data. In \eqref{eq:obj-q-gen-mix}, the state posterior $p(s_t| \ubm{x}, \bm{H}^{\mathrm{old}})$ is computed by forward-backward algorithm. The posterior of $\kappa$ is:
\begin{align}\label{eq:kappa-posterior}
  p(\kappa| s, \ubm{x}; \bm{H}^{\mathrm{old}})
  &=  \frac{p(\kappa, \ubm{x}| s; \bm{H}^{\mathrm{old}})}{p(\ubm{x}| s,\bm{H}^{\mathrm{old}})} \nonumber \\
  & = \frac{\pi_{s, \kappa}^{\mathrm{old}} p(\bm{x}| s, \kappa, \bm{H}^{\mathrm{old}})}{\sum_{\kappa=1}^{K}  \pi_{s, \kappa}^{\mathrm{old}} p(\bm{x}| s, \kappa,\bm{H}^{\mathrm{old}})},
\end{align}
where the last equation is due to the fact that $\bm{x}_t$ among sequence $\ubm{x}$ depends on $s_t, \kappa_t$ only.

By substituting \eqref{eq:changel-variable} and \eqref{eq:cat-jacobian} into \eqref{eq:obj-q-gen-mix}, we have cost function for neural networks as
\begin{align}\label{eq:obj-q-gen-mix}
  &\Qq(\bm{\Theta}; \bm{H}^{\mathrm{old}}) \nonumber \\
  =& \frac{1}{R}\hspace{-3pt}\sum_{r=1}^{R}\hspace{-3pt} \sum_{t=1}^{{T}^{r}}\hspace{-3pt} \sum_{\smtr{s}{t}{r}=1}^{|\Ss|} \hspace{-3pt} \sum_{\smtr{\kappa}{t}{r}=1}^{K}p(\smtr{s}{t}{r}| \ubmr{x}{r}; \bm{H}^{\mathrm{old}})p(\smtr{\kappa}{t}{r}|\smtr{s}{t}{r}, \ubmr{x}{r}; \bm{H}^{\mathrm{old}}) \nonumber\\
  &\left[ \log\, p_{\smtr{s}{t}{r}, \smtr{\kappa}{t}{r}}(\bm{f}_{\smtr{s}{t}{r}, \smtr{\kappa}{t}{r}}(\bmtr{x}{t}{r})) + \sum_{l=1}^{L}\log\,| \det (\nabla{\bm{f}_{s,\kappa}^{[l]}})|\right].
\end{align}
The generators of GenHMM can simply use standard Gaussian distribution for latent variables $\bm{z} \sim p_{s,\kappa}(\bm{z})$. It is straightforward to employ batch-size stochastic gradient decent to maximize $\Qq(\bm{\Theta; \bm{H}^{\mathrm{old}}})$ w.r.t. parameters of generators, i.e. $\bm{\Phi}$, since the dataset represented by $\hat{p}(\ubm{x})$ can be too large to do whole-dataset iterations.

@@

$p(\smtr{s}{1}{r};\bm{H})$ is the probability of initial state of GenHMM for $r$-th sequential sample. Actually $q_i = p({s}_{1} =i) $, $i= 1, 2, \cdots, |\Ss|$ for $\bm{H}$. Solution to problem:
\begin{align}
  \bm{q} &= \uargmax{\bm{q}} \Qq(\bm{q}; \bm{H}^{\mathrm{old}}), \nonumber \\
         &\mathrm{s.t.} \sum_{i=1}^{ |\Ss| }q_i = 1, q_i \geq 0, \forall s.
\end{align}
is
\begin{equation}\label{eq:update-initial-state-prob}
  q_i = \frac{1}{R} \sum_{r=1}^{R} p(\smtr{s}{1}{r}=i | \ubmr{x}{r}; \bm{H}^{\mathrm{old}}), \forall\; i = 1, 2, \cdots, |\Ss|.
\end{equation}

@@

\subsection{Algorithm of GenHMM and Convergence Analysis}

@@

\subsection{Experimental Setups}
The dataset used for sequential data modeling and classification is TIMIT dataset where the speech is sampled at $16$kHz. For extraction of feature vectors, we use $25$ms frame length and $10$ms frame shift to convert sound track into standard Mel-frequency cepstral coefficients (MFCCs) features. The deltas and delta-deltas features are also concatenated and used in our experiments.

@@

In all the following experiments, all generators $\left\{\bm{g}_{s,\kappa}|s\in \Ss, \kappa=1, 2, \cdots, K \right\}$ of GenHMM are implemented as flow models. To be more specific, our generator structure follows the structure of flow model in RealNVP \cite{2016arXiv160508803D}. The coupling layer shown as \eqref{eq-gl} maps a part of its input signal identically. The following coupling layer would alternate the input signal older such that after two consequential coupling layers, there is no signal remains the same any more. We term such two consequential coupling layers as a \textit{flow block}. In our experiments, a generator $\bm{g}_{s,\kappa}$ consists of four flow blocks. The latent source signal is assumed as standard Gaussian, i.e. $p_{s,\kappa}(\bm{z})$ is density function of standard Gaussian.

In learning of GenHMM, the number of states is set as $|\Ss| \in \left\{ 3,4,5 \right\}$. The exact number of states for a phoneme is decided by computing the average length of MFCC frames per phone in training dataset, and clipping the average length into $\left\{ 3,4,5 \right\}$. Transition matrix $\bm{A}$ is initialized as upper triangular matrix for each GenHMM.

>>>>

Another track of related work is hybrid method of temporal neural network models and HMM. In \cite{liu2019lstmHmmHyb,buys2018bridging,vik2016rnnHmm}, a long short-term memory (LSTM) model/recurrent neural network (RNN) is combined with HMM as hybrid. A hierarchical training is carried out by: i) training a HMM first, ii) and then doing modified training of LSTM using trained HMM. This hierarchical training procedure is motivated by the intuition of using LSTM or RNN to fill in the gap where HMM can not learn.

@@

In this work, we propose a generative model based HMM. Specifically, we propose a generator-mixed HMM, termed as GenHMM, where generators are realized by neural networks to help the model gain high modeling capacity.
Our proposed model, GenHMM:
\begin{itemize}
  \item firstly has high modeling capacity of sequential data, due to the neural network based generators;
\item has advantage of easy to do training. Training of GenHMM employs expectation maximization (EM) framework. Training a GenHMM is as easy as training a GMM-HMM model, while configuration of GenHMM is flexible;
% \item has high modeling capability of sequence data, though its training and using are handy
\item is able to compute loglikelihood exactly and efficiently.
\end{itemize}
Instead of using softmax for probability representation, our GenHMM has tractability of exact loglikelihood of given sequential data, which is based on change of variable formula. To make the loglikelihood computation efficient, neural network based generators of GenHMM are implemented as flow models.

Our contributions in the paper are as follows:
\begin{itemize}
\item Proposing a neural network based HMM for sequential data modeling, i.e. GenHMM. GenHMM has the tractability of exact likelihood.
\item Designing practical algorithm for training GenHMM under EM framework. Stochastic gradient  search in batch fashion is embedded in this algorithm.
\item Giving convergence analysis for GenHMM under the proposed learning algorithm.
\item Verifying the efficiency of the proposed model on practical sequential data.
\end{itemize}

\section{Generator-mixed HMM (GenHMM)}

@@

Our framework is a Hidden Markov Model (HMM). A HMM model $\bm{H}$ defined in a hypothesis space $\Hh$, i.e. $\bm{H} \in \Hh$, is capable to model time-span signal $\ubar{\bm{x}} = \left[ \bm{x}_1, \cdots, \bm{x}_T\right]^{\intercal}$, where $\bm{x}_t\in \RR^{N}$ is the signal at time $t$, $[\cdot]^{\intercal}$ denotes transpose, and $T$ denotes the time length\footnote{The length for difference sequential data can be different.}. Using HMM for signal representation is illustrated in Figure~\ref{fig:hmm}. The model assumption is that different instant signal of $\ubar{\bm{x}}$ is generated by a different signal source associated with a hidden state of HMM, while the source selection is done by a discrete hidden markov process. We define the hypothesis set of HMM as $\Hh := \{\bm{H} | \{\Ss, \bm{q}, \bm{A}, p(\bm{x}|{s}; \bm{\Phi}_{s})\}$, where:
\begin{itemize}
\item $\Ss$ is the set of hidden states of HMM $\bm{H}$.
\item $\bm{q} = \left[ q_1, q_2, \cdots, q_{|\Ss|}\right]^\intercal$ is initial distribution of HMM $\bm{H}$ with $|\Ss|$ as cardinality of $\Ss$. For a state $s$, $q_s = p(s;\bm{H})$. In the following of this paper, we use $s_t$ to denote the state $s$ at time $t$.
\item $\bm{A}$ is the transition matrix for the HMM $\bm{H}$ of size $|\Ss| \times |\Ss|$ and $\bm{A}_{i,j} = p(s_{t+1}=j|s_{t}=i; \bm{H})$.
\item For given hidden state $s$, the density function of observable signal is $p({\bm{x}}|{s};\bm{\Phi}_{s})$, where $\bm{\Phi}_{s}$ is the parameter set that defines this probabilistic model. Denote $\bm{\Phi} = \left\{ \bm{\Phi}_{s}| s \in \Ss \right\}$.
\end{itemize}

In the framework of HMM, at each time instance, signal $\bm{x}_t$ is assumed to be generated by a distribution with density function $p(\bm{x}_t| s_t; \bm{\Phi}_{s_t})$, and $s_t$ is decided by the hidden markov process. Putting these together gives us the probabilistic model $p(\ubm{x};\bm{H})$.

\subsection{Generative Model of GenHMM}

@@

In this section, we introduce a neural network based generator-mixed HMM model. We term it as GenHMM. GenHMM has a hidden markov process as a typical HMM does, while the probabilistic model of GenHMM for each hidden state is induced by mixture of neural network based generators. For given state $s\in \Ss$, we define its probabilistic model in GenHMM as
\begin{equation}
  p(\bm{x}| s; \bm{\Phi}_{s}) = \sum_{\kappa=1}^{K}\pi_{s, \kappa} p(\bm{x}| s, \kappa; \bm{\theta}_{s, \kappa}),
\end{equation}
where $\kappa$ is a random variable following a categorical distribution, with probability $\pi_{s, \kappa} = p(\kappa | s; \bm{H})$. Naturally $\sum_{\kappa = 1}^{K} \pi_{s, \kappa}= 1$. Denote $\bm{\pi}_{s} = [\pi_{s,1}, \pi_{s,2}, \cdots, \pi_{s,K}]^{\intercal}$,
$\bm{\theta}_s = \left\{ \bm{\theta}_{s, \kappa}| \kappa = 1, 2, \cdots, K \right\}$. 
We define $p(\bm{x}| s, \kappa; \bm{\theta}_{s, \kappa})$ as induced distribution by a generator $\bm{g}_{s,\kappa}: \RR^{N}\rightarrow\RR^{N}$, such that $\bm{x}=\bm{g}_{s, \kappa}(\bm{z})$, where $\bm{z}$ is a latent variable following a distribution with density function $p_{s,\kappa}(\bm{z})$. Assuming $\bm{g}_{s, \kappa}$ is invertible, by change of variable, we have
\begin{equation}\label{eq:changel-variable}
  p(\bm{x}| s, \kappa; \bm{\theta}_{s, \kappa}) = p_{s,\kappa}(\bm{z})\bigg| \det\left( \pd{\bm{g}_{s,\kappa}(\bm{z})}{\bm{z}} \right)\bigg|^{-1}.
\end{equation}

For a state $s$ of GenHMM, the induced probability distribution has the signal flow illustration as shown in Figure~\ref{fig:gen-mix}, in which the generator identity is up to the random variable $\kappa$.

@@

The first problem that we need to address is to deal with the hidden sequential variable $\ubm{s}$ and $\ubm{\kappa}$  in model $\bm{H}$, where $\ubm{s}=[ \bm{s}_1, \bm{s}_2, \cdots, \bm{s}_T ]^{\intercal}$ and $\ubm{\kappa} = [\bm{\kappa}_1, \bm{\kappa}_2, \cdots, \bm{\kappa}_T]^{\intercal}$. For a sequential observable variable $\ubm{x}$, $\ubm{s}$ is the hidden state sequence corresponding to $\ubm{x}$, and $\ubm{\kappa}$ is the hidden variable sequence representing the generator identity sequence that actually generates $\ubm{x}$.

@@

In this section, we detail the solution for realizing and learning GenHMM. % We begin with introducing flow model that we are going to use to model our generators $\{\bm{g}_{s, \kappa}| s\in\Ss, \kappa=1,2,\cdots, K\}$ for efficient likelihood computation. The learning method of HMM is followed then. 
The convergence of GenHMM is also discussed in this section.

\subsection{Realizing $\bm{g}_{s,\kappa}$ by a Flow Model}
We use a feed-forward neural network to implement each generator $\bm{g}_{s,\kappa}$.
Define  $\bm{g}_{s,\kappa}$ as a $L$-layer neural network and we formulate its mapping by layer-wise concatenation:
$\bm{g}_{s,\kappa}=\bm{g}_{s,\kappa}^{[L]}\circ \bm{g}_{s,\kappa}^{[L-1]}\circ \cdots
\circ \bm{g}_{s,\kappa}^{[1]}$, where superscipt $[l]$ denotes the $l$'s layer index and $\circ$ denotes mapping concatenation. Assume $\bm{g}_{s,\kappa}$ is invertible and denotes its inverse mapping as $\bm{f}_{s,\kappa}=\bm{g}_{s,\kappa}^{-1}$. For a latent variable $\bm{z}$ with density function $p_{s,\kappa}(\bm{z})$, the generated signal $\bm{x}$ follows an induced distribution with density function \eqref{eq:changel-variable}. We illustrate the signal flow between latent variable $\bm{z}$ and observable variable $\bm{x}$ as follows:
\begin{equation*}
  \vspace{-8pt}
  \centering
  \begin{tikzpicture}
    \node (z) at (0,0) {};
    \node at ($(z)-(0.5,0)$){$\bm{z}=\bm{h}_0$};
    \node (xi1) at (1.5,0) {$\bm{h}_1$};
    \node (xi2) at (3,0) {};
    \node (xi3) at (4.5,0){};
    \node (x) at (6,0) {};
    \node at ($(x)+(0.5,0)$){${\bm{h}_L=\bm{x}}$};
    \draw[->] ($(z) + (0.3,0.1)$) -- node[above]{$\bm{g}_{s,\kappa}^{[1]}$} ($(xi1)+(-0.3,0.1)$); 
    \draw[->] ($(xi1)-(0.3,0.1)$) -- node[below]{${\bm{f}}_{s,\kappa}^{[1]}$}($(z) - (-0.3,0.1)$);
    \draw[->] ($(xi1) + (0.3,0.1)$) -- node[above]{$\bm{g}_{s,\kappa}^{[2]}$} ($(xi2)+(-0.3,0.1)$); 
    \draw[->] ($(xi2)-(0.3,0.1)$) -- node[below]{${\bm{f}}_{s,\kappa}^{[2]}$}($(xi1) - (-0.3,0.1)$);
    \draw[->] ($(xi3) + (0.3,0.1)$) -- node[above]{$\bm{g}_{s,\kappa}^{[L]}$} ($(x)+(-0.3,0.1)$); 
    \draw[->] ($(x)-(0.3,0.1)$) -- node[below]{${\bm{f}}_{s,\kappa}^{[L]}$}($(xi3) - (-0.3,0.1)$);
    \draw[dotted,line width = 0.3 mm] (xi2) -- (xi3);
  \end{tikzpicture},
\end{equation*}
where $\bm{f}_{s,\kappa}^{[l]}$ is the $l$-th layer of $\bm{f}_{s,\kappa}$. We have $\bm{z}=\bm{f}_{s,\kappa}(\bm{x})$. If every layer of $\bm{g}_{s,\kappa}$ is invertible,
the full feed-forward neural network is invertible. Flow model, proposed in \cite{DBLP:journals/corr/DinhKB14} as an image generating model, is such a invertible
feed-forward neural network. It is further improved in subsquential works \cite{2016arXiv160508803D,kingma2018glow} for high-fidelity and high-resolution image generating and representation. Flow model also has advantages as efficient Jacobian computation and low computational complexity. %Computation of inverse mapping for flow-based network and determinant of Jacobin were addressed in the literature.

@@

We firstly address the generator learning problem, i.e. $\umax{\bm{\Theta}} \Qq(\bm{\Theta}; \bm{H}^{\mathrm{old}})$. This is boiled down to maximizing the cost function of neural networks that can be formulated as
\begin{align}\label{eq:obj-q-gen-mix-log}
  &\Qq(\bm{\Theta}; \bm{H}^{\mathrm{old}}) \nonumber \\
  = &\frac{1}{R}\sum_{r=1}^{R}\sum_{\ubmr{s}{r}}\sum_{\ubmr{\kappa}{r}}{p(\ubmr{s}{r}, \ubmr{\kappa}{r}| \ubmr{x}{r}; \bm{H}^{\mathrm{old}})} \sum_{t=1}^{{T}^{r}}\log\,p(\bmtr{x}{t}{r} | \smtr{s}{t}{r}, \smtr{\kappa}{t}{r}; \bm{H}) \nonumber \\
  =& \frac{1}{R}\sum_{r=1}^{R} \sum_{t=1}^{{T}^{r}} \sum_{\smtr{s}{t}{r}=1}^{|\Ss|}  \sum_{\smtr{\kappa}{t}{r}=1}^{K}p(\smtr{s}{t}{r}| \ubmr{x}{r}; \bm{H}^{\mathrm{old}})p(\smtr{\kappa}{t}{r}|\smtr{s}{t}{r}, \ubmr{x}{r}; \bm{H}^{\mathrm{old}}) \nonumber\\
  &  \log\, p(\bmtr{x}{t}{r} | \smtr{s}{t}{r}, \smtr{\kappa}{t}{r}; \bm{H}), 
\end{align}
where $T^r$ is the length of the $r$-th sequential data. In \eqref{eq:obj-q-gen-mix-log}, the state posterior $p(s_t| \ubm{x}, \bm{H}^{\mathrm{old}})$ is computed by forward-backward algorithm. The posterior of $\kappa$ is:
\begin{align}\label{eq:kappa-posterior}
  p(\kappa| s, \ubm{x}; \bm{H}^{\mathrm{old}})
  &=  \frac{p(\kappa, \ubm{x}| s; \bm{H}^{\mathrm{old}})}{p(\ubm{x}| s,\bm{H}^{\mathrm{old}})} \nonumber \\
  & = \frac{\pi_{s, \kappa}^{\mathrm{old}} p(\bm{x}| s, \kappa, \bm{H}^{\mathrm{old}})}{\sum_{\kappa=1}^{K}  \pi_{s, \kappa}^{\mathrm{old}} p(\bm{x}| s, \kappa,\bm{H}^{\mathrm{old}})},
\end{align}
where the last equation is due to the fact that $\bm{x}_t$ among sequence $\ubm{x}$ depends on $s_t, \kappa_t$ only.

By substituting \eqref{eq:changel-variable} and \eqref{eq:cat-jacobian} into \eqref{eq:obj-q-gen-mix-log}, we have cost function for neural networks as
\begin{align}\label{eq:obj-q-gen-mix}
  &\Qq(\bm{\Theta}; \bm{H}^{\mathrm{old}}) \nonumber \\
  =& \frac{1}{R}\hspace{-3pt}\sum_{r=1}^{R}\hspace{-3pt} \sum_{t=1}^{{T}^{r}}\hspace{-3pt} \sum_{\smtr{s}{t}{r}=1}^{|\Ss|} \hspace{-3pt} \sum_{\smtr{\kappa}{t}{r}=1}^{K}p(\smtr{s}{t}{r}| \ubmr{x}{r}; \bm{H}^{\mathrm{old}})p(\smtr{\kappa}{t}{r}|\smtr{s}{t}{r}, \ubmr{x}{r}; \bm{H}^{\mathrm{old}}) \nonumber\\
  &\left[ \log\, p_{\smtr{s}{t}{r}, \smtr{\kappa}{t}{r}}(\bm{f}_{\smtr{s}{t}{r}, \smtr{\kappa}{t}{r}}(\bmtr{x}{t}{r})) + \sum_{l=1}^{L}\log\,| \det (\nabla{\bm{f}_{s,\kappa}^{[l]}})|\right].
\end{align}
The generators of GenHMM can simply use standard Gaussian distribution for latent variables $\bm{z} \sim p_{s,\kappa}(\bm{z})$. It is straightforward to employ batch-size stochastic gradient decent to maximize $\Qq(\bm{\Theta; \bm{H}^{\mathrm{old}}})$ w.r.t. parameters of generators, i.e. $\bm{\Phi}$, since the dataset represented by $\hat{p}(\ubm{x})$ can be too large to do whole-dataset iterations.

@@

$p(\smtr{s}{1}{r};\bm{H})$ is the probability of initial state of GenHMM for $r$-th sequential sample. Actually $q_i = p({s}_{1} =i;\bm{H}) $, $i= 1, 2, \cdots, |\Ss|$. Solution to problem:
\begin{align}
  \bm{q} &= \uargmax{\bm{q}} \Qq(\bm{q}; \bm{H}^{\mathrm{old}}), \nonumber \\
         &\mathrm{s.t.} \sum_{i=1}^{ |\Ss| }q_i = 1, q_i \geq 0, \forall s.
\end{align}
is
\begin{equation}\label{eq:update-initial-state-prob}
  q_i = \frac{1}{R} \sum_{r=1}^{R} p(\smtr{s}{1}{r}=i | \ubmr{x}{r}; \bm{H}^{\mathrm{old}}), \forall\; i = 1, 2, \cdots, |\Ss|.
\end{equation}

@@

\subsection{Algorithm of GenHMM}

@@

\subsection{Experimental Setups}
The dataset used for sequential data modeling and classification is TIMIT dataset where the speech signal is sampled at $16$kHz. For extraction of feature vectors, we use $25$ms frame length and $10$ms frame shift to convert sound track into standard Mel-frequency cepstral coefficients (MFCCs) features. The deltas and delta-deltas features are also concatenated and used in our experiments.

@@

In all the following experiments, all generators $\left\{\bm{g}_{s,\kappa}|s\in \Ss, \kappa=1, 2, \cdots, K \right\}$ of GenHMM are implemented as flow models. To be more specific, our generator structure follows the structure of flow model in RealNVP \cite{2016arXiv160508803D}. The coupling layer shown as \eqref{eq-gl} maps a part of its input signal identically. The following coupling layer would alternate the input signal order such that after two consequential coupling layers, there is no signal remains the same any more. We term such two consequential coupling layers as a \textit{flow block}. In our experiments, a generator $\bm{g}_{s,\kappa}$ consists of four flow blocks. The latent source signal is assumed as standard Gaussian, i.e. $p_{s,\kappa}(\bm{z})$ is density function of standard Gaussian.

In learning of GenHMM, the number of states is set as $|\Ss| \in \left\{ 3,4,5 \right\}$. The exact number of states for a phoneme is decided by computing the average length of MFCC frames per phone in training dataset, and clipping the average length into $\left\{ 3,4,5 \right\}$. Transition matrix $\bm{A}$ is initialized as upper triangular matrix for GenHMM.
